---
title: "MATH 306: Proposal"
output: html_document
date: "2025-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Proposal

### A brief description of the dataset 

### Data visualisations

### A brief description of the topic that you are proposing to work with

### A brief description of the application that you are proposing to work with

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(mosaic)
library(ggplot2)
```
### Preparing dataset


```{r data-visualisations}

startups <- read.csv("startup_data.csv")

colnames(startups) <- c("startup_name", "industry", "funding_rounds", "funding_amount", "valuation", "revenue", "employees", "market_share_percent", "profitable", "year_founded", "region", "exit_status")

colnames(startups)

```


## Data Visualisations



```{r}
ggplot(startups, aes(x = funding_amount, y = valuation, color = industry)) +
  geom_point() +
  scale_color_viridis_d() +
  labs(
    title = "Funding Amount vs. Valuation by Industry",
    x = "Funding Amount (M USD)",
    y = "Valuation (M USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```




This plot shows the relationship between funding and valuation, key metrics for startup success. By separating the data by industry, we can see which industries are more likely to attract higher funding and achieve higher valuations.



```{r}
ggplot(startups, aes(x = region, y = market_share_percent, fill = region)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(
    title = "Market Share Distribution by Region",
    x = "Region",
    y = "Market Share (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


This boxplot visualizes the distribution of market share across different regions. It helps to compare the spread of market share within each region and helps us see which regions have the highest and lowest market share.




```{r}

ggplot(startups, aes(x = employees , y = revenue)) +
  geom_point(aes(color = exit_status), size = 3) +
  scale_color_viridis_d() +
  facet_wrap(~exit_status) +
  labs(
    title = "Employees vs. Revenue by Exit Status",
    x = "Employees",
    y = "Revenue (M USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```



This faceted scatter plot shows how employee count relates to revenue within each exit status category. By faceting the data, we can examine whether startups of different exit statuses tend to have different relationships between these two variables. This could reveal trends, such as whether large, high-revenue startups are more likely to be acquired or go public.

