---
title: "MATH 306: Proposal"
author: "Noah Jarbeau, Phattiya Matrakul, Jelizaveta Tsalaja"
output: "github_document"
date: "2025-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Proposal

### **Topic:** Start-Up Growth and Financial Trend Prediction Using Least Squares Classification 

Startups operate in highly competitive environments, where factors such as funding, valuation, and market share are important in determining their long-term success. Our study aims to analyze startup growth and financial trends using Least Squares Classification to predict a startup’s exit status (Private, IPO, or Acquired). The exit status is a key indicator of a startup’s success as it can reflect how well each start-up company has scaled, secured funding, and sustained profitability. By applying Least Squares Classification (LSC), we aim to model the relationship between key startup features and their likelihood of achieving a successful 


### About Data

The Startup Growth & Funding [1] dataset we are using includes information on 500 startups across multiple industries, relating to financial, operational, and market-related characteristics. It contains 12 variables by mixing numerical and categorical features that contribute to an understanding of startup performance. 

The numerical variables include funding rounds, funding amount (in millions of USD), valuation (in millions of USD), revenue (in millions of USD), number of employees, market share percentage, and the year the startup was founded. These variables provide insight into the financial and operational scale of the startups. Funding rounds tell us how many rounds a company goes through in seeking funds. To provide some contexts, a higher number of rounds means there are more investors. Valuation is “the analytical process of determining the current (or projected) worth of an asset or company” [2]. It can consider future growth potential, market trends and competition. Revenue is the amount of money the company earns before expenses are deducted. Market share is “the percent of total sales in an industry generated by a particular company” [3].

Categorical variables include startup name, industry (e.g., AI, FinTech, HealthTech), profitability (binary indicator: profitable or not), exit status (IPO, acquired, or private), and region (e.g., Europe, South America, North America). We hope that this dataset provides a diverse set of variables, both numerical and categorical, that could be used for classification tasks aimed at understanding startup performance and outcomes.

### Why This Data is Interesting

This dataset is interesting because it provides key indicators of startup success, allowing us to  explore funding trends, industry-specific growth patterns, and financial sustainability. Startups vary significantly in how they secure funding, achieve profitability, and scale their operations, so it is important to identify the factors that contribute to their outcome. By applying Least Squares Classification, we can analyze how variables such as funding, market share, and profitability status influence whether a startup remains private, goes public through an IPO, or is acquired by a larger company.

### Questions This Data Can Help Address

- What financial and operational factors are most predictive of a startup achieving IPO or acquisition?

- Do startups with higher funding and valuation have a greater likelihood of successful exit outcomes?

- How do industry-specific differences impact startup growth and profitability?



```{r message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(mosaic)
library(ggplot2)
```


### Preparing dataset


```{r data-visualisations}

startups <- read.csv("startup_data.csv")

colnames(startups) <- c("startup_name", "industry", "funding_rounds", "funding_amount", "valuation", "revenue", "employees", "market_share_percent", "profitable", "year_founded", "region", "exit_status")

colnames(startups)

```


### Data Visualisations


#### Funding amount vs. Valuation by Industry

```{r}
ggplot(startups, aes(x = funding_amount, y = valuation, color = industry)) +
  geom_point() +
  scale_color_viridis_d() +
  labs(
    title = "Funding Amount vs. Valuation by Industry",
    x = "Funding Amount (M USD)",
    y = "Valuation (M USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


This plot shows the relationship between funding and valuation, key metrics for startup success. By separating the data by industry, we can see which industries are more likely to attract higher funding and achieve higher valuations.


#### Market Share Distribution by Region


```{r}
ggplot(startups, aes(x = region, y = market_share_percent, fill = region)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(
    title = "Market Share Distribution by Region",
    x = "Region",
    y = "Market Share (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


This boxplot visualizes the distribution of market share across different regions. It helps to compare the spread of market share within each region and helps us see which regions have the highest and lowest market share.


#### Employees vs. Revenue by Exit Status


```{r}

ggplot(startups, aes(x = employees , y = revenue)) +
  geom_point(aes(color = exit_status), size = 3) +
  scale_color_viridis_d() +
  facet_wrap(~exit_status) +
  labs(
    title = "Employees vs. Revenue by Exit Status",
    x = "Employees",
    y = "Revenue (M USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```



This faceted scatter plot shows how employee count relates to revenue within each exit status category. By faceting the data, we can examine whether startups of different exit statuses tend to have different relationships between these two variables. This could reveal trends, such as whether large, high-revenue startups are more likely to be acquired or go public.


### Appendix 1

Our group's second choice dataset is https://www.kaggle.com/datasets/mohammedarfathr/smartwatch-health-data-uncleaned/data [4]. We are hoping to use Least Squares Classification (LSC) to classify smartwatch users based on various health metrics and behaviors such as heart rate, blood oxygen level, step count, sleep duration, activity level, and stress level. Our would be to predict a user's activity level (categorized as Sedentary, Active, or Highly Active) based on the features collected from their smartwatch. The chosen dataset includes data from smartwatch users. The population of interest in this dataset is individuals who monitor their health using a smartwatch, focusing on users' physiological metrics such as heart rate, blood oxygen level and step count. Our chosen dataset consists of data collected from 10,000 smartwatch users. There are both numerical and categorical variables in the dataset. The numerical variables (Heart Rate, Blood Oxygen Level, Step Count, Sleep Duration, and Stress Level) have different units (beats per minute, percentage, and hours) and ranges. The Activity Level variable, which we hope to predict with our classification, categorizes users into three groups based on their physical activity, while the Stress Level is an ordinal variable indicating varying degrees of perceived stress.

```{r}
smartwatch <- read.csv("unclean_smartwatch_health_data.csv")
head(smartwatch)
```



### Appendix 2

Our group considered the following data sets, in addition to our first and second choices:

- Heart-disease: https://archive.ics.uci.edu/dataset/45/heart+disease [5]

- Poisonous Mushroom: https://archive.ics.uci.edu/dataset/73/mushroom [6]

- Breast Cancer: https://archive.ics.uci.edu/dataset/15/breast+cancer+wisconsin+original [7]

- 2023 season MLB Results https://github.com/fivethirtyeight/data/tree/master/mlb-elo [8]

### References

[1]  Ashar, Samay. “Startup Growth & Funding Trends.” Kaggle.com, 2025. https://doi.org/10853721/aa2f631868a1247c4a97ae6fb7063149.

[2]Hayes, Adam. “Market Share: What It Is and the Formula for Calculating It.” Investopedia, 23 Aug. 2024, www.investopedia.com/terms/m/marketshare.asp.

[3] Chen, James. “How the Valuation Process Works.” Investopedia, 30 June 2023, www.investopedia.com/terms/v/valuation.asp.

[4] Arfath, Mohammed . “Smartwatch Health Data (Uncleaned).” Kaggle.com, 2025. https://www.kaggle.com/datasets/mohammedarfathr/smartwatch-health-data-uncleaned/data.

[5]  Janosi, A., Steinbrunn, W., Pfisterer, M., & Detrano, R. (1989). Heart Disease [Dataset]. UCI Machine Learning Repository. https://doi.org/10.24432/C52P4X.

[6] Mushroom [Dataset]. (1981). UCI Machine Learning Repository. https://doi.org/10.24432/C5959T.

[7] Wolberg, W. (1990). Breast Cancer Wisconsin (Original) [Dataset]. UCI Machine Learning Repository. https://doi.org/10.24432/C5HP4Z.

[8]  Boice, Jay . “Data/Mlb-Elo at Master · Fivethirtyeight/Data.” GitHub, 2019, github.com/fivethirtyeight/data/tree/master/mlb-elo.
