---
title: "MATH 306: Proposal"
output: "github_document"
date: "2025-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Proposal

### A brief description of the dataset 

### Data visualisations

### A brief description of the topic that you are proposing to work with

### A brief description of the application that you are proposing to work with

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(mosaic)
library(ggplot2)
```


## Data

The dataset that we chose contains information on 500 startups, focusing on various characteristics such as industry, funding, valuation, revenue, market share, profitability, and geographic region. The population of interest is startups across different sectors, and the data includes 12 variables. Among these, numerical variables include funding rounds, funding amount (in millions of USD), valuation (in millions of USD), revenue (in millions of USD), number of employees, market share percentage, and the year the startup was founded. These variables provide insight into the financial and operational scale of the startups. Categorical variables include startup name, industry (e.g., AI, FinTech, HealthTech), profitability (binary indicator: profitable or not), exit status (IPO, acquired, or private), and region (e.g., Europe, South America, North America). We hope that this dataset provides a diverse set of variables, both numerical and categorical, that could be used for classification tasks aimed at understanding startup performance and outcomes.


### Preparing dataset


```{r data-visualisations}

startups <- read.csv("startup_data.csv")

colnames(startups) <- c("startup_name", "industry", "funding_rounds", "funding_amount", "valuation", "revenue", "employees", "market_share_percent", "profitable", "year_founded", "region", "exit_status")

colnames(startups)

```


## Data Visualisations


#### Funding amount vs. Valuation by Industry

```{r}
ggplot(startups, aes(x = funding_amount, y = valuation, color = industry)) +
  geom_point() +
  scale_color_viridis_d() +
  labs(
    title = "Funding Amount vs. Valuation by Industry",
    x = "Funding Amount (M USD)",
    y = "Valuation (M USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```




This plot shows the relationship between funding and valuation, key metrics for startup success. By separating the data by industry, we can see which industries are more likely to attract higher funding and achieve higher valuations.



#### Market Share Distribution by Region


```{r}
ggplot(startups, aes(x = region, y = market_share_percent, fill = region)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(
    title = "Market Share Distribution by Region",
    x = "Region",
    y = "Market Share (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


This boxplot visualizes the distribution of market share across different regions. It helps to compare the spread of market share within each region and helps us see which regions have the highest and lowest market share.


#### Employees vs. Revenue by Exit Status


```{r}

ggplot(startups, aes(x = employees , y = revenue)) +
  geom_point(aes(color = exit_status), size = 3) +
  scale_color_viridis_d() +
  facet_wrap(~exit_status) +
  labs(
    title = "Employees vs. Revenue by Exit Status",
    x = "Employees",
    y = "Revenue (M USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```



This faceted scatter plot shows how employee count relates to revenue within each exit status category. By faceting the data, we can examine whether startups of different exit statuses tend to have different relationships between these two variables. This could reveal trends, such as whether large, high-revenue startups are more likely to be acquired or go public.


## Appendix 1

Our group's second choice dataset is https://www.kaggle.com/datasets/mohammedarfathr/smartwatch-health-data-uncleaned/data. We are hoping to use Least Squares Classification (LSC) to classify smartwatch users based on various health metrics and behaviors such as heart rate, blood oxygen level, step count, sleep duration, activity level, and stress level. Our would be to predict a user's activity level (categorized as Sedentary, Active, or Highly Active) based on the features collected from their smartwatch. The chosen dataset includes data from smartwatch users. The population of interest in this dataset is individuals who monitor their health using a smartwatch, focusing on users' physiological metrics such as heart rate, blood oxygen level and step count. Our chosen dataset consists of data collected from 10,000 smartwatch users. There are both numerical and categorical variables in the dataset. The numerical variables (Heart Rate, Blood Oxygen Level, Step Count, Sleep Duration, and Stress Level) have different units (beats per minute, percentage, and hours) and ranges. The Activity Level variable, which we hope to predict with our classification, categorizes users into three groups based on their physical activity, while the Stress Level is an ordinal variable indicating varying degrees of perceived stress.

```{r}
smartwatch <- read.csv("unclean_smartwatch_health_data.csv")
head(smartwatch)
```



## Appendix 2

Our group considered the following data sets, in addition to our first and second choices:

- Heart-disease: https://archive.ics.uci.edu/dataset/45/heart+disease

- Poisonous Mushroom: https://archive.ics.uci.edu/dataset/73/mushroom 

- Breast Cancer: https://archive.ics.uci.edu/dataset/15/breast+cancer+wisconsin+original 

- 2023 season MLB Results https://github.com/fivethirtyeight/data/tree/master/mlb-elo