---
title: "MATH 306: Project Draft 1 Data Work"
author: "Noah Jarbeau, Phattiya Matrakul, Jelizaveta Tsalaja"
output: "github_document"
date: "2025-03-31"
---

### Graphs in the Proposal 


```{r message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(mosaic)
library(ggplot2)
```


```{r data-visualisations}

startups <- read.csv("/cloud/project/startup_data.csv")

colnames(startups) <- c("startup_name", "industry", "funding_rounds", "funding_amount", "valuation", "revenue", "employees", "market_share_percent", "profitable", "year_founded", "region", "exit_status")

colnames(startups)

```

```{r}
ggplot(startups, aes(x = funding_amount, y = valuation, color = industry)) +
  geom_point() +
  scale_color_viridis_d() +
  labs(
    title = "Funding Amount vs. Valuation by Industry",
    x = "Funding Amount (M USD)",
    y = "Valuation (M USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


This plot shows the relationship between funding and valuation, key metrics for startup success. By separating the data by industry, we can see which industries are more likely to attract higher funding and achieve higher valuations.

```{r}
ggplot(startups, aes(x = region, y = market_share_percent, fill = region)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(
    title = "Market Share Distribution by Region",
    x = "Region",
    y = "Market Share (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

This boxplot visualizes the distribution of market share across different regions. It helps to compare the spread of market share within each region and helps us see which regions have the highest and lowest market share.

```{r}

ggplot(startups, aes(x = employees , y = revenue)) +
  geom_point(aes(color = exit_status), size = 3) +
  scale_color_viridis_d() +
  facet_wrap(~exit_status) +
  labs(
    title = "Employees vs. Revenue by Exit Status",
    x = "Employees",
    y = "Revenue (M USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

This faceted scatter plot shows how employee count relates to revenue within each exit status category. By faceting the data, we can examine whether startups of different exit statuses tend to have different relationships between these two variables. This could reveal trends, such as whether large, high-revenue startups are more likely to be acquired or go public.


Now, that we have identified out main research question, we will create data visualisations that are more direclty related to it.


## Which financial and operational factors best predict a startup’s path to a successful exit?

### Funding Amount vs. Exit Status

Goal: To analyze whether higher funding amounts correlate with successful exits (Acquired, IPO)
Visualization: A boxplot of funding amounts grouped by exit status

```{r funding-vs-exit}
ggplot(startups, aes(x = exit_status, y = funding_amount, fill = exit_status)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_viridis_d() +
  labs(title = "Funding Amount vs. Exit Status",
       x = "Exit Status",
       y = "Funding Amount (log scale)") +
  theme_minimal()
```

### Employees vs. Exit Rate

Goal: To examine whether the number of employees is a predictor of a successful exit
Visualization: A violin plot showing the distribution of employees across exit statuses

```{r employees-exit}
ggplot(startups, aes(x = exit_status, y = employees, fill = exit_status)) +
  scale_fill_viridis_d() +
  geom_violin(alpha = 0.8) +
  scale_y_log10() +  # Log scale to handle large variation
  labs(title = "Employee Count vs. Exit Status",
       x = "Exit Status",
       y = "Number of Employees (log scale)") +
  theme_minimal()
```

### Year Founded vs. Exit Rate

Goal: To examine whether the year of the foundation is a predictor of a successful exit

```{r}
startups <- startups %>%
  mutate(year_group = cut(year_founded, 
                          breaks = seq(floor(min(year_founded) / 5) * 5, max(year_founded), by = 5), 
                          include.lowest = TRUE, 
                          right = FALSE, 
                          labels = paste(seq(floor(min(year_founded) / 5) * 5, max(year_founded) - 5, by = 5),
                                         seq(floor(min(year_founded) / 5) * 5 + 4, max(year_founded) - 1, by = 5), 
                                         sep = "-")))

ggplot(startups, aes(x = year_group, fill = exit_status)) +
  geom_bar(position = "fill", alpha = 0.7) +
  scale_fill_viridis_d(direction = -1) +
  labs(title = "Exit Rate Over Time (Grouped by 5-Year Intervals)",
       x = "Year Founded (5-Year Groups)",
       y = "Proportion of Exit Status") +
  theme_minimal()
```


### Exit Status by Profitability over time

```{r}
ggplot(startups, aes(x = year_group, fill = exit_status)) +
  geom_bar(position = "fill", alpha = 0.7) +
  scale_fill_viridis_d(direction = -1) +  
  facet_wrap(~ profitable, labeller = labeller(profitable = c("0" = "Not Profitable", "1" = "Profitable"))) +  
  labs(title = "Exit Rate Over Time (Grouped by 5-Year Intervals) - Split by Profitability",
       x = "Year Founded (5-Year Groups)",
       y = "Proportion of Exit Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Industry vs. Exit Rate

Goal: To see which industries have the highest rates of successful exits
Visualization: A bar plot showing the proportion of startups in each industry that had a successful exit

```{r}
ggplot(startups, aes(x = industry, fill = exit_status)) +
  scale_fill_viridis_d(direction = -1) +
  geom_bar(position = "fill", alpha = 0.7) +  # Stacked bar chart normalized to proportions
  labs(title = "Exit Rate by Industry",
       x = "Industry",
       y = "Proportion of Exit Status") +
  theme_minimal() +
  coord_flip() 
```

### Industry vs. Exit Rate by Region

Goal: To see which industries have the highest rates of successful exits and whether they vary based on the region

```{r}
ggplot(startups, aes(x = industry, fill = exit_status)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d(direction = -1, alpha = 0.7) +
  facet_wrap(~ region) +  # Split by region for deeper insight
  labs(title = "Exit Rate by Industry Across Regions",
       x = "Industry",
       y = "Proportion of Exit Status") +
  theme_minimal() +
  coord_flip()
```



### Exit Status by Profitability

```{r}
ggplot(startups, aes(x = factor(profitable), fill = exit_status)) +
  geom_bar(position = "fill", alpha = 0.7) +  
  scale_fill_viridis_d(direction = -1) +  
  labs(title = "Exit Status by Profitability",
       x = "Profitability (0 = Not Profitable, 1 = Profitable)",
       y = "Proportion of Exit Status") +
  theme_minimal()
```

### Funding Amount vs. Exit Status

```{r}
ggplot(startups, aes(x = exit_status, y = funding_amount, fill = exit_status)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_viridis_d() +
  labs(title = "Exit Status by Funding Amount",
       x = "Exit Status",
       y = "Funding Amount (in Millions)") +
  theme_minimal()
```


### Valuation vs. Exit Status

```{r}
ggplot(startups, aes(x = exit_status, y = valuation, fill = exit_status)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_viridis_d() +
  labs(title = "Exit Status by Valuation",
       x = "Exit Status",
       y = "Valuation (in Millions)") +
  theme_minimal()
```


### Market Share % vs. Exit Status

```{r}
ggplot(startups, aes(x = exit_status, y = market_share_percent, fill = exit_status)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_viridis_d() +
  labs(title = "Exit Status by Market Share Percentage",
       x = "Exit Status",
       y = "Market Share (%)") +
  theme_minimal()
```

